\section{Image Scanning Microscopy}

Modern biological and biomedical research frequently demands the ability to image
structures at the nanoscale — from individual proteins a few nanometres in size, to
organelles and whole cells at the micrometre scale. Optical microscopy has long been
the workhorse of such investigations, owing to its non-invasive nature and compatibility
with live specimens. Yet imaging at this scale is far from straightforward: two
fundamental requirements, \textit{high contrast} and \textit{high resolution}, are
each difficult to achieve and often stand in tension with one another.

\subsection{Fluorescence Microscopy and the Diffraction Limit}

Fluorescence microscopy addresses the contrast challenge by exploiting the phenomenon
of fluorescence: certain molecules (fluorophores) absorb excitation light at a shorter
wavelength and re-emit it at a longer, less energetic wavelength. By selectively
labelling specific cellular structures with fluorescent markers, one can visualise
components of interest against a dark background with high specificity, effectively
filtering out the rest of the sample. A standard fluorescence microscope routes a
laser excitation beam through an objective lens onto the specimen; the emitted
fluorescent light is then separated from the excitation by a dichroic mirror and
emission filter, and collected by a detector to form an image.

Resolution, however, is fundamentally bounded by the wave nature of light. A lens
cannot focus an electromagnetic beam to an infinitely small spot; instead, each
point source in the sample is imaged as a diffuse blob known as the \textit{Point
Spread Function} (PSF). The width of the PSF determines the smallest resolvable
feature: two emitters closer together than the PSF width are indistinguishable. This
is formalised by the \textit{Abbe diffraction limit}, which states that the minimal
resolvable distance $d$ satisfies
\begin{equation}
    d = \frac{\lambda}{2 \, \mathrm{NA}} = \frac{\lambda}{2 n \sin\theta}
    \approx 200\text{--}250 \; \text{nm},
\end{equation}
where $\lambda$ is the emission wavelength, $\mathrm{NA} = n\sin\theta$ is the
numerical aperture of the objective, $n$ is the refractive index of the immersion
medium, and $\theta$ is the half-angle of the acceptance cone.
Equivalently, in the frequency domain, the \textit{Optical Transfer Function} (OTF) —
the Fourier transform of the PSF — acts as a low-pass filter, irreversibly suppressing
spatial frequencies beyond a cutoff determined by the numerical aperture. The
consequence is that widefield fluorescence microscopy, despite its excellent contrast,
is resolution-limited to roughly a quarter of a micrometre, leaving many sub-cellular
structures inaccessible.

\subsection{Confocal Laser Scanning Microscopy}

The \textit{Confocal Laser Scanning Microscope} (CLSM) improves upon widefield
imaging by replacing camera-based wide-area detection with a point-by-point scanning
strategy. A focused laser beam is raster-scanned across the specimen using galvanometric
mirrors; at each scan position $\mathbf{x}_s$, the emitted fluorescence passes back
through the objective and is focused through a small \textit{pinhole} aperture onto a
single-element detector (typically a photodiode or avalanche photodetector), which
records a single photon-count value. The pinhole physically rejects out-of-focus
light, conferring two important advantages: optical sectioning (the ability to
reconstruct thin focal planes within a thick specimen) and, when the pinhole is closed
down, a modest improvement in lateral resolution beyond the Abbe limit by up to a
factor of $\sqrt{2}$.

These gains come at a cost. Closing the pinhole necessarily discards a large fraction
of the collected photons, severely reducing the signal-to-noise ratio (SNR). The
experimenter thus faces a hard trade-off: a wide pinhole recovers photon counts but
degrades both sectioning and resolution, while a narrow pinhole improves spatial
quality at the expense of a dim, noisy image. This photon-budget constraint is
particularly problematic in live-cell imaging, where fluorophore concentrations and
permissible laser powers are both limited by phototoxicity.

\subsection{Image Scanning Microscopy: Concept and Data Structure}

\textit{Image Scanning Microscopy} (ISM) resolves the CLSM trade-off by replacing the
single-element detector with a two-dimensional \textit{Single-Photon Avalanche Diode
(SPAD) array}. The array, subtending approximately $1.4$
Airy Units at the detector plane, is centred on the optical axis. For each scan
position $\mathbf{x}_s = (x_s, y_s)$, rather than integrating all arriving photons
into one number, every pixel $\mathbf{x}_d = (x_d, y_d)$ of the detector array
records its own photon count independently. A complete ISM acquisition therefore
produces a \textit{four-dimensional dataset} $i(\mathbf{x}_s \mid \mathbf{x}_d)$,
consisting of $(N_x \times N_y)$ micro-images (one per scan point), each of which is
a small image of the sample as seen by a particular detector element. Equivalently,
one may reorganise the same data as $N_\mathrm{ch}$ scanned images (one per detector
pixel), each representing a full field of view acquired from a different vantage
point on the detector plane. This high-dimensional dataset carries substantially more
spatial information than a conventional confocal acquisition and is a defining feature
of the ISM modality.

\subsection{Forward Model and the Inverse Problem}

The image formation process in fluorescence microscopy is well described as a \textit{linear shift-invariant} (LSI) system. The observed image is the convolution of the unknown fluorophore distribution $\mathbf{u} \in \mathbb{R}^N_+$ with the system's PSF $h$, corrupted by Poisson photon-counting noise $\mathcal{P}$: \begin{equation} \mathbf{y} = \mathcal{P}(\mathbf{h} * \mathbf{u}) = \mathcal{P}(\mathbf{A}\mathbf{u}), \end{equation} where $\mathbf{A} \in \mathbb{R}^{N \times N}$ is the Toeplitz matrix encoding convolution with the PSF. In CLSM the PSF is a product of the excitation PSF and the effective detection PSF shaped by the pinhole; in the ISM setting, the forward model generalises to a family of $N_\mathrm{ch}$ equations \begin{equation} \mathbf{y}_d = \mathcal{P}(\mathbf{h}_d * \mathbf{u}) = \mathcal{P}(\mathbf{A}_d \mathbf{u}), \qquad d = 1, \ldots, N_\mathrm{ch}, \end{equation} where each detector element $\mathbf{x}_d$ produces a scanned image governed by a \textit{distinct} PSF $h_d(\mathbf{x}_s \mid \mathbf{x}_d) = h_\mathrm{exc}(\mathbf{x}_s) \times h_\mathrm{det}(\mathbf{x}_s - \mathbf{x}_d)$. The effective PSF for detector element $d$ can be approximated as a shifted copy of a common PSF, $h(\mathbf{x}_s \mid \mathbf{x}_d) \approx h(\mathbf{x}_s - \boldsymbol{\mu}(\mathbf{x}_d))$, where $\boldsymbol{\mu}(\mathbf{x}_d)$ is the \textit{shift-vector} associated with that detector element — computed as the shift maximising cross-correlation between micro-images. Because each detector element occupies a different position relative to the optical axis, the $N_\mathrm{ch}$ scanned images represent shifted, hence \textit{redundant}, views of the same underlying object. This redundancy is the key resource that ISM exploits to surpass the confocal resolution limit while recovering the photon budget lost by closing the pinhole. The SPAD array introduces an additional practical advantage: it produces no electronic read-out noise, so photon statistics are governed purely by Poisson shot noise — a clean statistical model amenable to principled computational treatment.

Recovering the fluorophore distribution $\mathbf{u}$ from the measured ISM dataset $\{i(\mathbf{x}_s \mid \mathbf{x}_d)\}$ is thus an \textit{inverse problem}: given the forward operators $\{\mathbf{A}_d\}$ and the noisy observations, infer the latent image. Naive approaches such as pixel-reassignment (summing the shifted micro-images) provide a fast but suboptimal estimate; modern reconstruction pipelines exploit the full 4D dataset through variational regularisation or learned priors to achieve simultaneous super-resolution and optical sectioning.

\subsection{From Microscopy to Tomography}
The inverse problem structure of ISM shares deep mathematical parallels with that of Computed Tomography, discussed in the following section. In both cases, one seeks to recover an unknown spatial distribution — a fluorophore map $\mathbf{u}(\mathbf{x})$ in ISM, an attenuation map $\mu(\mathbf{x})$ in CT — from a collection of indirect, noisy measurements related to it through a linear forward operator. In ISM the forward operator is a convolution (a consequence of the LSI imaging model), while in CT it is the Radon transform (a consequence of straight-line X-ray propagation); yet both are compact linear operators whose inversion is ill-posed and requires regularisation. Moreover, just as ISM acquires multiple shifted views of the same object — each providing complementary spatial information — CT acquires projections from multiple angles, with each angular view illuminating a different set of line integrals through the target. The transition from microscopy to tomography is therefore not merely a change of application domain, but a shift in the geometry of the forward operator: from local blurring in the spatial domain to global integration along lines. Both settings exemplify the broader programme of computational imaging, in which the design of the acquisition, the modelling of the forward physics, and the inversion algorithm are co-optimised to extract the maximum information about an object from measurements that are always finite, incomplete, and noisy.